Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-31 doc/papers/cgo2014/paper.tex
Add OpenMP part.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{subcaption}

@@

C/C++ have been at the hearth of high performance computing for many years.
Compiler infrasturctures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provide 
good abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.

@@

@@

\subsection{Vectorization and Parallelization}

\cite{esterie2012boost, pyhpc2013, wpmvp2014}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{todo}

@@

C/C++ have been at the hearth of high performance computing for many years.
Compiler infrasturctures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
good abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.

@@

Note that turning such Numpy expressions into expression templates is
equivalent to converting from a strict evaluation to a non-strict, or lazy,
evaluation. This is always legal in the contect of Numpy because none of the
computations considered for expression template conversion have side effect.

@@

\subsection{Parallelization and Vectorization}

A simplified version of the evaluation loop for an expression template is
illustrated on Listing~\ref{lst:eval-et}. The loop over each dimension of the
target array expression \texttt{to} is made explicit through a recursive
template call. The computation of the origin expression is only triggered for
the innermost loop.

By construction of the expression template, the \texttt{eval} method used in
\texttt{from.eval(i)} does not have side effects, and the Python to C++
compiler can assert there is no shared memory between \texttt{to} and any
component of \texttt{from}. In that case, all loops can be made parallel, and
the innermost loop is eventually vectorizable.

\begin{lstlisting}[language=c++, label={lst:eval-et}, caption={Evaluation body of an expression template}]
template<size_t N> struct D {};

// real evaluation loop
template<class A, class E>
void eval(A& to, E const& from, D<1>)
{
  size_t n = to.shape[A::ndims-1];
  for(size_t i = 0; i < n; ++i)
    to[i] = from.eval(i);
}

// recursive evaluation loops
template<class A, class E, size_t Dim>
void eval(A& to, E const& from, D<Dim>)
{
  size_t n = to.shape[A::ndims-Dim];
  for(size_t i = 0; i < n; ++i)
    eval(to[i], from[i], D<Dim-1>);
}
\end{lstlisting}

\subsubsection{Parallelisation of Expression templates}

A naive way to parallelize the parameteric loop nest from
Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
OpenMP~\cite{openmp4} standard. Depending on the number of dimension of the
input array, this may lead to nested parallel regions. When OpenMP nested
parallel regions are activated, each parallel region spaws as many thread as
possible, which leads to oversubscription and generally bad performances. When
disabled, only the first parallel region spawns threads, which is likely to be
inefficient depending on the input array shape, e.g. $(1,100000)$. Moreover,
even if they do not start new threads, each nested parallel region implies a
costly overhead. The \texttt{collapse} clause is not compatible with the
parametric loop version.

A first approach would be to use recursieve tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}

A conservative approach is to only start a parallel region if the loop trip
count is large enough. We avoid by using the \texttt{if} clause for parallel
regions to do so in order to prevent the unnecessary overhead. Instead the loop
statement is duplicated and guarded by the appropriate check, where only one
loop holds the parallel annotations.

Note that this parallelization scheme is compatible with the manual
parallelization of Python programs proposed in~\cite{pyhpc2013}.

\subsubsection{Vectorization of Expression Templates}

A first exploration of vectorization of Python constructs is done in~\cite{wpmvp2014}.
